---
title: "Coevolution with a seed bank"
authors: "Jay T. Lennon & Daniel A Schwartz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Analyze composition of mutations from pooled population sequencing

# Setup Work Environment
```{r}
# Load dependencies
library(here)
library(tidyverse)
require("vegan")

```

# Load data
```{r}
mutdat <- read_csv(here("data/mult_host.csv")) %>% 
  rename(trt = 1)
```

# PCoA procedures
```{r}

# Define treatments and data
seed <- str_detect(mutdat$trt, "long")
phage <- str_detect(mutdat$trt, "SPO1")
# seed <- c(1,1,1,1,1,1,0,0,0,0,0,0)
# phage <- c(0,0,0,1,1,1,0,0,0,1,1,1)
mut <- cbind(seed, phage, mutdat[,2:ncol(mutdat)])

# "manhattan", "euclidean", "canberra", "clark", "bray", "kulczynski", "jaccard", "gower", "altGower", "morisita", "horn", "mountford", "raup", "binomial", "chao", "cao", "mahalanobis", "chisq" or "chord".

# Calculate pairwise distances
mut.dist <- vegdist(mut, method = "bray", binary = "FALSE")

# Principal Coordinates Analysis (PCoA)
pc <- cmdscale(mut.dist, eig = TRUE, k = 3)
explainvar1 <- round(pc$eig[1] / sum(pc$eig), 3) * 100
explainvar2 <- round(pc$eig[2] / sum(pc$eig), 3) * 100
explainvar3 <- round(pc$eig[3] / sum(pc$eig), 3) * 100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)
##  two first PCoAs explain >100% variation



```

Consulting with WRS, he noticed we have a negative eigenvalue on the order of the second largest positive eigenvalue, which would be considered a large negative eigenvalue. 

```{r}
qplot(1:12,pc$eig)+ theme_classic()
```

WRS suggested we try a PCA, since prcomp uses singular value decomposition instead of eigendecomposition on the covariance matrix, so it doesn’t return negative.

## PCA Plot
```{r}
pca <- prcomp(mut.dist)

var_explained <- pca$sdev^2/sum(pca$sdev^2)

p <- cbind(mut[,1:2],pca$x) %>% 
  as.data.frame %>%
    mutate(seed = if_else(seed==1, "with seed bank", "no seed bank"),
         phage= if_else(phage==1, "with phage", "no phage") ) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color = seed, shape = phage), size=4, stroke=1)+
  # geom_polygon(data = dl, linetype = 3 ,fill="transparent",
  #              aes(x=x, y =y, group = interaction(seed, phage),color = seed))+
  theme_bw(base_size=32) + 
  labs(x=paste0("PC1: ",round(var_explained[1]*100,1),"%"),
       y=paste0("PC2: ",round(var_explained[2]*100,1),"%")) +
  geom_hline(yintercept = 0, linetype = 3)+
  geom_vline(xintercept = 0, linetype = 3)+
  scale_shape_manual(values = c(21,22))+
  scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-1,1)) +
  scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-1,1))+
  theme_classic(base_size = 16)

p  
ggsave(here("analysis","PCA_gg.png"),p, width = 5, height = 3)
```

# Elbow plot

How many PCs to include in stats?
```{r}

sum.pca <- summary(pca)

prop_var <- sum.pca$importance[2,]
plot(1:12,prop_var, type = "b", xlab = "PC", ylab = "Proportion of Variance")

cum_var <- sum.pca$importance[3,]
plot(1:12,cum_var, type = "b", xlab = "PC", ylab = "Cumulative Proportion")
```
First to PCs explain `r  sum.pca$importance[3,2]*100`% of the variartion

```{r}
# PERMANOVA on PCA
seed <- mut[,1]
phage <- mut[,2]
perm <- adonis2(pca$x[,c(1:2)] ~ seed * phage, method = "euclidean", binary = FALSE)
perm

```

In summary, we analyze the composition of genes under selection by PCA analysis on multiplicity data . PERMANOVA on the first two axis (that together explain >98% of variation) shows that treatments significantly alter the composition, as does the treatment interaction. The F value shows that the seed-bank has the strongest effect, and that the interaction is much weaker, but all are significant

# Ellipses

The ggplot function of stat_ellipse does not allow CI ellipses on less than 4 data points. We have three points per treatment. However three points should be alowed "because your CI depends on the variance, which takes two degrees of freedom".  

According to stat_ellipses help "The method for calculating the ellipses has been modified from car::dataEllipse (Fox and Weisberg, 2011)". The limit on 3 points does not exist in the original function.

```{r}

library(car)
d.ellipse <- cbind(mut[,1:2],pca$x) %>% 
  as.data.frame %>%
  mutate(seed = if_else(seed, "with seed bank", "no seed bank"),
         phage= if_else(phage, "with phage", "no phage"),
         grp=interaction(seed,phage))

el <- dataEllipse(d.ellipse$PC1, d.ellipse$PC2, groups = d.ellipse$grp)


# unpack list
dl <- rbind(
cbind("no seed bank.no phage",el$`no seed bank.no phage`$`0.95`),
cbind("with seed bank.no phage",el$`with seed bank.no phage`$`0.95`),
cbind("no seed bank.with phage",el$`no seed bank.with phage`$`0.95`),
cbind("with seed bank.with phage",el$`with seed bank.with phage`$`0.95`)
)


dl <- dl %>% 
  as_tibble() %>% 
  mutate(x= as.numeric(x), y=as.numeric(y)) %>% 
  separate(V1, into = c("seed", "phage"),remove = F, sep = "\\.")

```

# PCA with ellipses

```{r}
p <- cbind(mut[,1:2],pca$x) %>% 
  as.data.frame %>%
    mutate(seed = if_else(seed==1, "with seed bank", "no seed bank"),
         phage= if_else(phage==1, "with phage", "no phage") ) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
    geom_polygon(data = dl, linetype = 2 ,fill="transparent", size =  1,
               aes(x=x, y =y, group = interaction(seed, phage), color = seed))+
  geom_point(aes(color = seed, shape = phage), size=4, stroke=1)+

  theme_bw(base_size=32) + 
  labs(x=paste0("PC1: ",round(var_explained[1]*100,1),"%"),
       y=paste0("PC2: ",round(var_explained[2]*100,1),"%")) +
  geom_hline(yintercept = 0, linetype = 3)+
  geom_vline(xintercept = 0, linetype = 3)+
  scale_shape_manual(values = c(21,22))+
  scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-1,1)) +
  scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-1,1))+
  theme_classic(base_size = 16)

p  

```

