---
title: "Coevolution with a seed bank"
authors: "Jay T. Lennon & Daniel A Schwartz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Analyze composition of mutations from pooled population sequencing

# Setup Work Environment

```{r}
# Load dependencies
library(here)
library(tidyverse)
library(vegan)
library(BiodiversityR)

```

# Load data

Matrix of multiplicity data organized as population X gene.

```{r}
mutdat <- read_csv(here("data/mult_host.csv")) %>% 
  rename(trt = 1)
```

# PCoA procedures

```{r}

# Define treatments and data
seed <- str_detect(mutdat$trt, "long")
phage <- str_detect(mutdat$trt, "SPO1")

# multiplicity data only
mut <- mutdat %>% select(-trt)


# Calculate pairwise distances
mut.dist <- vegdist(mut, method = "bray", binary = "FALSE")


# Principal Coordinates Analysis (PCoA)
pc <- cmdscale(mut.dist, eig = TRUE, k = nrow(mut)-1)
explainvar1 <- round(pc$eig[1] / sum(pc$eig), 3) * 100
explainvar2 <- round(pc$eig[2] / sum(pc$eig), 3) * 100
explainvar3 <- round(pc$eig[3] / sum(pc$eig), 3) * 100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)



p <-  
  as_tibble(pc$points, .name_repair = "universal" ) %>% 
  rename_with(~gsub("...", "PCoA", .x, fixed = TRUE)) %>% 
  mutate(seed = seed, phage = phage) %>% 
  relocate(seed,phage,.before = 1) %>% 
    mutate(seed = if_else(seed==1, "with seed bank", "no seed bank"),
         phage= if_else(phage==1, "with phage", "no phage") ) %>% 
  ggplot(aes(x=PCoA1,y=PCoA2)) + 
  geom_point(aes(color = seed, fill = seed, shape = phage), size=3, stroke=1,alpha=0.8)+
  # geom_polygon(data = dl, linetype = 3 ,fill="transparent",
  #              aes(x=x, y =y, group = interaction(seed, phage),color = seed))+
  theme_bw(base_size=32) + 
  labs(x=paste0("PCo 1 (",round(explainvar1,1),"%)"),
       y=paste0("PCo 2 (",round(explainvar2,1),"%)")) +
  geom_hline(yintercept = 0, linetype = 3)+
  geom_vline(xintercept = 0, linetype = 3)+
  scale_shape_manual(values = c(21,24))+
  scale_fill_grey(end = 0.8)+
  scale_color_grey(end = 0.6)+
  scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-0.4,0.4)) +
  scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
                     limits = c(-0.4,0.4))+
  theme_classic(base_size = 16)

ggsave(here("analysis","PCoA_mult_host.png"),p, width = 5, height = 3)

p



```

No negative eigenvalues

```{r}
qplot(1:12,pc$eig, geom = "col")+ 
  theme_classic()+
  scale_x_continuous(breaks = 1:12)+
  xlab("PCo")
```
How many PCs to include in stats?

```{r}
plot(1:length(pc$eig),pc$eig,type = "b", 
     xlab = "PC", ylab = "Eigenvalue")


prop_var <- round(pc$eig / sum(pc$eig), 3) * 100
plot(1:length(pc$eig),prop_var, type = "b", 
     xlab = "PC", ylab = "Proportion of Variance")

# PCs explaining 90% variation
pc_var90 <- min(which(cumsum(prop_var)>90))
plot(1:length(pc$eig),cumsum(prop_var), type = "b", 
     xlab = "PC", ylab = "Cumulative Proportion", ylim = c(0,100))
abline(v=pc_var90,h=90, col = c("grey", "blue"))




```

First `r pc_var90` PCs explain >90% of the variartion


# PERMANOVA

```{r}
# # on multiplicity data
# perm <-
#   adonis2(mut.dist ~ seed * phage,
#           binary = FALSE, permutations = 9999)

# on PCs explaining >90% var
perm <-
  adonis2(pc$points[,1:pc_var90] ~ seed * phage,method = "euclidean",
          binary = FALSE, permutations = 9999)

perm
```


# Gene correlations

```{r}
# genes in delta6 ---------------------------------
delta6_168 <- read_csv(here("data/teichoic_acid","delta6_168_cds_matched.csv"),trim_ws = T, name_repair = "universal")
# categories_168 <- read_csv(here("data/teichoic_acid","geneCategories-2022-06-27.csv"),trim_ws = T, name_repair = "universal")
SW.export_168 <- read_csv(here("data/teichoic_acid","subtiwiki.gene.export.2022-06-27.csv"),trim_ws = T)
d.genes <- left_join(delta6_168,SW.export_168 , by = c("locus_tag.168"="locus"))


# Test correlation to PCoA axes ---------------------------------------
gene.corr <- add.spec.scores(pc, mut, method = "cor.scores")$cproj  
gene.corr <- 
  tibble(gene = rownames(gene.corr)) %>% 
  bind_cols(as_tibble(gene.corr ))

# Genes correlated with PCo1 --------------------------------

fit <- envfit(pc, mut, choices=1, perm = 999)

d.fit1 <- tibble(gene = names(fit$vectors$r),
       r = fit$vectors$r,
       pvals = fit$vectors$pvals)

# combine with Correlation for significant genes
sig_genes1 <- d.fit1 %>% 
  filter(pvals<0.05) %>% 
  left_join(., gene.corr %>% select(gene, cor = Dim1)) 

# add annotations
sig_genes1 <- sig_genes1 %>% 
  left_join(., d.genes, by = c("gene" = "locus_tag.d6")) 

# export positively correlated
sig_genes1 %>% 
  filter(cor > 0) %>% 
  arrange(desc(abs(cor))) %>% 
  select(locus_tag.d6=gene, locus_tag.168, title, description, `function`, strand, cor,   P_value = pvals) %>% 
  write_csv(here("data", "significant_genes_pc1_positive.csv"))

# export negatively correlated
sig_genes1 %>% 
  filter(cor < 0) %>% 
  arrange(desc(abs(cor))) %>% 
  select(locus_tag.d6=gene, locus_tag.168, title, description, `function`, strand, cor,   P_value = pvals) %>% 
  write_csv(here("data", "significant_genes_pc1_negative.csv"))


# Genes correlated wit PCo2 ---------------------------------------------
fit <- envfit(pc, mut, choices=2, perm = 999)

d.fit2 <- tibble(gene = names(fit$vectors$r),
       r = fit$vectors$r,
       pvals = fit$vectors$pvals)

# Correlation of significant genes
sig_genes2 <- d.fit2 %>% 
  filter(pvals<0.05) %>% 
  left_join(., gene.corr %>% select(gene, cor = Dim2)) 

# add annotations
sig_genes2 <- sig_genes2 %>% 
  left_join(., d.genes, by = c("gene" = "locus_tag.d6")) 

# export positively correlated
sig_genes2 %>% 
  filter(cor > 0) %>% 
  arrange(desc(abs(cor))) %>% 
  select(locus_tag.d6=gene, locus_tag.168, title, description, `function`, strand, cor,   P_value = pvals) %>% 
  write_csv(here("data", "significant_genes_pc2_positive.csv"))

# export negatively correlated
sig_genes2 %>% 
  filter(cor < 0) %>% 
  arrange(desc(abs(cor))) %>% 
  select(locus_tag.d6=gene, locus_tag.168, title, description, `function`, strand, cor,  P_value = pvals) %>% 
  write_csv(here("data", "significant_genes_pc2_negative.csv"))

```

# Ellipses

The ggplot function of stat_ellipse does not allow CI ellipses on less than 4 data points. We have three points per treatment. However three points should be alowed "because your CI depends on the variance, which takes two degrees of freedom".

According to stat_ellipses help "The method for calculating the ellipses has been modified from car::dataEllipse (Fox and Weisberg, 2011)". The limit on 3 points does not exist in the original function.

```{r}

library(car)
d.ellipse <-  cbind(mut[,1:2],pc$points) %>% 
  as.data.frame %>%
  mutate(seed = if_else(seed, "with seed bank", "no seed bank"),
         phage= if_else(phage, "with phage", "no phage"),
         grp=interaction(seed,phage))

el <- dataEllipse(d.ellipse$`1`, d.ellipse$`2`, groups = d.ellipse$grp)


# unpack list
dl <- rbind(
cbind("no seed bank.no phage",el$`no seed bank.no phage`$`0.95`),
cbind("with seed bank.no phage",el$`with seed bank.no phage`$`0.95`),
cbind("no seed bank.with phage",el$`no seed bank.with phage`$`0.95`),
cbind("with seed bank.with phage",el$`with seed bank.with phage`$`0.95`)
)


dl <- dl %>% 
  as_tibble() %>% 
  mutate(x= as.numeric(x), y=as.numeric(y)) %>% 
  separate(V1, into = c("seed", "phage"),remove = F, sep = "\\.")

```

# PCA with ellipses

```{r}
p <-  
  as_tibble(pc$points, .name_repair = "universal" ) %>% 
  rename_with(~gsub("...", "PCoA", .x, fixed = TRUE)) %>% 
  mutate(seed = seed, phage = phage) %>% 
  relocate(seed,phage,.before = 1) %>% 
    mutate(seed = if_else(seed==1, "with seed bank", "no seed bank"),
         phage= if_else(phage==1, "with phage", "no phage") ) %>% 
  ggplot(aes(x=PCoA1,y=PCoA2)) + 
  geom_point(aes(color = seed, fill = seed, shape = phage), size=3, stroke=1,alpha=0.8)+
  geom_polygon(data = dl, linetype = 3 ,fill="transparent",
               aes(x=x, y =y, group = interaction(seed, phage),color = seed))+
  theme_bw(base_size=32) + 
  labs(x=paste0("PCo 1 (",round(explainvar1,1),"%)"),
       y=paste0("PCo 2 (",round(explainvar2,1),"%)")) +
  geom_hline(yintercept = 0, linetype = 3)+
  geom_vline(xintercept = 0, linetype = 3)+
  scale_shape_manual(values = c(21,24))+
  scale_fill_grey(end = 0.8)+
  scale_color_grey(end = 0.6)+
  # scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
  #                    limits = c(-0.4,0.4)) +
  # scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL),
  #                    limits = c(-0.4,0.4))+
  theme_classic(base_size = 16)

# ggsave(here("analysis","PCA_hellinger2.png"),p, width = 5, height = 3)

p  

```
