---
title: "Distribution of host mutatations"
author: "Daniel Schwartz"
output: html_document
---

```{r setup, include=FALSE}
# Load dependencies
library(here)
library(tidyverse)
library(cowplot)
require("vegan")
library(scales)

knitr::opts_chunk$set(echo = TRUE)
```

## load data

```{r load data}
mutdat <- read_csv(here("data/mult_host.csv")) %>% 
  rename("trt" = 1)
```

## Heatmap of matrix

```{r}
mut_mat <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  pivot_wider(names_from = trt, values_from = m) %>% 
  select(-gene) %>% 
  as.matrix()

heatmap(as.matrix(mut_mat),
        margins = c(13, 0))
```
## Observed richness

### presence absence

```{r}

mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > 0, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()

```

### by multiplicity
```{r}

mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  group_by(trt) %>% 
  summarise(sum.m = sum(m)) %>% 
  arrange(sum.m) %>% 
  ggplot(aes(trt, sum.m)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()

```
They all add up to 1, the data has been normalized.

## Rank-abundance

```{r}

RAC <- function(x = ""){
x = as.vector(x)
x.ab = x[x > 0]
x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
return(x.ab.ranked)
}

d.rac <- tibble()
for(i in 1:nrow(mutdat)){
  
 d.rac <- 
  tibble(m = RAC(mutdat[i,-1]), 
         rank = 1:length(m),
         trt = mutdat$trt[i]) %>% 
    bind_rows(d.rac, .)

  
}


# KS test for differences between distributions, done by Will in Python code
d.ks <- 
  tibble(phage = c("with phage", "no phage"),
         # seed.bank=NA, pop = NA,
         d.stat = c(expression(italic("D")~"= 0.4073"),
                    expression(italic("D")~"= 0.4117")),
         lab = expression(italic("P")~"< 0.0001"),
         # lab = expression(italic("P")~"<"~10^{-4}),
         x= 200, y = 1e-1)

p <- d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with seed bank", "no seed bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with phage", "no phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
           
  ggplot(aes(x= rank, y = m)) + 
  geom_line(aes(group = interaction(seed.bank,pop), color = seed.bank), size = 0.8)+
  geom_text(data = d.ks, aes(label = d.stat,x = x, y=y), parse = T, hjust = 0)+
  geom_text(data = d.ks, aes(label = lab,x = x, y=y/2), parse = T, hjust = 0)+
  facet_wrap(~phage)+
  theme_classic(base_size = 18)+
  panel_border(color = "black", size = 1.5)+
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), 
                # limits = c(1e-5, NA)
                )+
  scale_color_grey(start = 0, end = 0.7)+
  annotation_logticks(sides = "l")+
  ylab("relative gene multiplicity")+
  theme(legend.position = c(0.8, 0.14),
      legend.title = element_blank(),
      legend.background = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"))

ggsave2(here("analysis","RAC_host-bySeedBank.png"), p, width = 7, height = 4)

p
```

```{r}
d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
           
  ggplot(aes(x= rank, y = m)) + 
  geom_line(aes(linetype = pop, color = phage), size =1)+
  theme_classic()+
  facet_wrap(~seed.bank)+
  panel_border(color = "black")+
  scale_y_log10()

p <- d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with seed bank", "no seed bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with phage", "no phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
           
  ggplot(aes(x= rank, y = m,  group = interaction(seed.bank,pop))) + 
  geom_line(aes(color = phage), size = 0.8)+
  facet_wrap(~seed.bank)+
  theme_classic(base_size = 18)+
  panel_border(color = "black", size = 1.5)+
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), 
                # limits = c(1e-5, NA)
                )+
  scale_color_grey(start = 0, end = 0.7)+
  annotation_logticks(sides = "l")+
  ylab("relative gene multiplicity")+
  theme(legend.position = c(0.8, 0.14),
      legend.title = element_blank(),
      legend.background = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"))

ggsave2(here("analysis","RAC_host_byPhage.png"), p, width = 7, height = 4)

p
```

# Model fitting to RAC

```{r}

x <- (mut_mat * 1) %>% t()
# mode(x) <- "integer"
x <- as.data.frame(x)

mod.radfit <- tibble()

for(i in 1:nrow(x)){
  trt <- rownames(x)[i]

  RACresults<- radfit(x[i,], family= Gamma) 
  
  
  mod.radfit <-  bind_rows(
            RACresults$models$Null[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Preemption[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Lognorma[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Zipf[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Mandelbrot[c("model","aic","rank", "coefficients")] %>% as_tibble(),
            ) %>% 
    mutate(trt = trt) %>% 
    bind_rows(mod.radfit, .)
  
  plot(RACresults, main = trt)
  
}


mod.radfit %>% 
  
  select(-coefficients) %>% 
  distinct() %>% 
  group_by(trt) %>% 
  arrange(aic) %>% 
  mutate(model.rank = row_number()) %>% 

  ggplot(aes(model, aic)) +
  geom_col(aes(fill = as.character(model.rank))) +
  facet_wrap(~trt)+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "bottom")
  
```

In all the best fitting model is the log-normal.

from QB:
"The expected abundance (a) of species at rank r is exp[log(µ) + log(σ)Φ], where Φ is a standard normal function, µ is the fitted mean abundance, and σ is the fitted standard deviation of abundances."


## compare fitted log-normal paraeters
 
 
```{r}
log.radfit <- mod.radfit %>% 
  filter(model == "Log-Normal") %>% 
  mutate(param = rep(c("log.mu", "log.sigma" ),12)) %>% 
  pivot_wider(names_from = param, values_from = coefficients) %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) 

print("==== Two way ANOVA for log MU parameter ====")
  mu.aov2 <- aov(log.mu ~ phage * seed.bank, data = log.radfit)
summary(mu.aov2)

print("==== Two way ANOVA for log SIGMA parameter ====")
  sigma.aov2 <- aov(log.sigma ~ phage * seed.bank, data = log.radfit)
summary(sigma.aov2)
```


```{r}
d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
  ggplot(aes(m)) + 
  geom_density(aes(fill  = seed.bank), alpha = 0.5)+
  theme_classic()+
  facet_wrap(phage~pop, nrow = 2)+
  panel_border(color = "black")+
  scale_x_log10()
```

## repeat PA on high M

```{r}
cutoff = 1e-4
p1 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-3
p2 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-2
p3 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-1
p4 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

plot_grid(p1,p2, p3, p4)
```

```{r}
top_genes <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  group_by(trt) %>% 
  arrange(desc(m), .by_group = T) %>% 
  slice_head(n = 10) %>% 
  pivot_wider(names_from = trt, values_from = m, values_fill = 0)



heatmap(as.matrix(top_genes %>% select(-gene)),
        margins = c(13, 0))
```

```{r}
top_genes <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  group_by(trt) %>% 
  arrange(desc(m), .by_group = T) %>% 
  slice_head(n = 4) %>% 
  pivot_wider(names_from = trt, values_from = m, values_fill = 0)



heatmap(as.matrix(top_genes %>% select(-gene)),
        margins = c(13, 0))
```