---
title: "Richness"
author: "Daniel Schwartz"
date: "2022-08-01"
output: html_document
---

```{r setup, include=FALSE}
# Load dependencies
library(here)
library(tidyverse)
library(cowplot)
require("vegan")


knitr::opts_chunk$set(echo = TRUE)
```

## load data

```{r load data}
mutdat <- read_csv(here("data/mult_host.csv")) %>% 
  rename("trt" = 1)
```

## Heatmap of matrix

```{r}
mut_mat <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  pivot_wider(names_from = trt, values_from = m) %>% 
  select(-gene) %>% 
  as.matrix()

heatmap(as.matrix(mut_mat),
        margins = c(13, 0))
```
## Observed richness

### presence absence

```{r}

mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > 0, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()

```

### by multiplicity
```{r}

mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  group_by(trt) %>% 
  summarise(sum.m = sum(m)) %>% 
  arrange(sum.m) %>% 
  ggplot(aes(trt, sum.m)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()

```
They all add up to 1, the data has been normalized.

## Rank-abundance

```{r}

RAC <- function(x = ""){
x = as.vector(x)
x.ab = x[x > 0]
x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
return(x.ab.ranked)
}

d.rac <- tibble()
for(i in 1:nrow(mutdat)){
  
 d.rac <- 
  tibble(m = RAC(mutdat[i,-1]), 
         rank = 1:length(m),
         trt = mutdat$trt[i]) %>% 
    bind_rows(d.rac, .)

  
}

d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
           
  ggplot(aes(x= rank, y = m)) + 
  geom_line(aes(linetype = pop, color = seed.bank))+
  theme_classic()+
  facet_wrap(~phage)+
  panel_border(color = "black")+
  scale_y_log10()
```

# Model fitting to RAC

```{r}

x <- (mut_mat * 1e7) %>% t()
mode(x) <- "integer"
x <- as.data.frame(x)

mod.radfit <- tibble()

for(i in 1:nrow(x)){
  trt <- rownames(x)[i]

  RACresults<- radfit(x[i,]) 
  
  
  mod.radfit <-  bind_rows(
            RACresults$models$Null[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Preemption[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Lognorma[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Zipf[c("model","aic","rank", "coefficients")] %>% as_tibble(),
 RACresults$models$Mandelbrot[c("model","aic","rank", "coefficients")] %>% as_tibble(),
            ) %>% 
    mutate(trt = trt) %>% 
    bind_rows(mod.radfit, .)
  
  plot(RACresults, main = trt)
  
}


mod.radfit %>% 
  
  select(-coefficients) %>% 
  distinct() %>% 
  group_by(trt) %>% 
  arrange(aic) %>% 
  mutate(model.rank = row_number()) %>% 

  ggplot(aes(model, aic)) +
  geom_col(aes(fill = as.character(model.rank))) +
  facet_wrap(~trt)+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "bottom")
  
```

In all except 2 samples the best fitting model is the log-normal.

from QB:
"The expected abundance (a) of species at rank r is exp[log(µ) + log(σ)Φ], where Φ is a standard normal function, µ is the fitted mean abundance, and σ is the fitted standard deviation of abundances."


## compare fitted log-normal paraeters
 
 
```{r}
log.radfit <- mod.radfit %>% 
  filter(model == "Log-Normal") %>% 
  mutate(param = rep(c("log.mu", "log.sigma" ),12)) %>% 
  pivot_wider(names_from = param, values_from = coefficients) %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) 

print("==== Two way ANOVA for log MU parameter ====")
  mu.aov2 <- aov(log.mu ~ phage * seed.bank, data = log.radfit)
summary(mu.aov2)

print("==== Two way ANOVA for log SIGMA parameter ====")
  sigma.aov2 <- aov(log.sigma ~ phage * seed.bank, data = log.radfit)
summary(sigma.aov2)
```


```{r}
d.rac %>% 
  mutate(seed.bank = if_else(str_detect(trt, "long"), "with-seed-bank", "no-seed_bank")) %>% 
  mutate(phage = if_else(str_detect(trt, "SPO1"), "with-phage", "no-phage")) %>% 
  mutate(pop = str_remove(trt, ".*_")) %>% 
  ggplot(aes(m)) + 
  geom_histogram(aes(fill  = seed.bank), bins = 50)+
  theme_classic()+
  facet_wrap(seed.bank~phage+pop, nrow = 2)+
  panel_border(color = "black")+
  scale_x_log10()
```

## repeat PA on high M

```{r}
cutoff = 1e-4
p1 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-3
p2 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-2
p3 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

cutoff = 1e-1
p4 <- mutdat %>% 
  pivot_longer(!trt, names_to = "gene", values_to = "m") %>% 
  mutate(pa = if_else(m > cutoff, 1,0)) %>% 
  group_by(trt) %>% 
  summarise(N.pa = sum(pa)) %>% 
  arrange(N.pa) %>% 
  ggplot(aes(trt, N.pa)) +
  geom_col()+
  coord_flip()+
  scale_x_discrete(limits=rev)+
  theme_classic()+
  ggtitle(paste("cut off =", cutoff))

plot_grid(p1,p2, p3, p4)
```

